---
title: "ITS Basic Bioinformatics Overview"
author: "PennCHOP Microbiome Program"
date: \today
geometry: margin=3cm
output: 
    pdf_document:
        template: forKnitting/toc_after.tex
        keep_tex: false
        toc: true
        toc_depth: 3
        includes:
            in_header: forKnitting/TeX_packages_commands.sty

---

\tableofcontents

<!-- ================================================================================================ -->
<!--   Beginning of Preamble : Preamble seldom requires change                                        -->
<!-- ================================================================================================ -->


<!-- knitr setup -->
```{r knitr setup, echo=FALSE}
### ================
###   knitr setup
### ================
library(knitr)
opts_chunk$set(
  tidy=FALSE,
  cache=FALSE,
  echo=FALSE,
  warning=FALSE,
  message=FALSE,
  dpi=100,
  fig.width=6,
  fig.height=5,
  fig.align = "center"
  )
```

<!-- R packages -->
```{r R packages, message=FALSE}
### ================
###   R packages
### ================
##install.packages(c("dplyr", "qiimer", "pander", "ape", "vegan", "ggplot2", "gplots", "pheatmap", "tidyr", "usedist", "readr", "tibble", "lme4", "magrittr"))
#This package will also help us more easily manipulate our data
library(reshape)
library(dplyr)
library(magrittr)
library(qiimer)
library(pander)
#Analyses of Phylogenetics and Evolution package. Required for tree calculations to be used with phyloseq
library(ape)
#The vegan package provides tools for descriptive community ecology. It has most basic functions of diversity analysis, community ordination and dissimilarity analysis. In general, this package is used for Bray-Curtis and Jaccard analyses.
library(vegan)
#Graphing package used in phyloseq. To edit the default setting of a plot, you need to use functions in this package.
library(ggplot2)
#This package is used to calculate and plot Venn diagrams as well as heatmaps
library(gplots)
library(pheatmap)
#This package will help us more easily manipulate our data, which are matrices
library(tidyr)
library(usedist)
library(readr)
library(tibble)
#Linear mixed-effects models like repeated measures analysis
library(lme4)
library(RColorBrewer)
library(here)
#used to read in mothur-formatted files
#library(phangorn)
#The phyloseq package seeks to address issues with multiple microbiome analysis packages by providing a set of functions that internally manage the organizing, linking, storing, and analyzing of phylogenetic sequencing data. In general, this package is used for UniFrac analyses.
#library(phyloseq)
#Pretty Venn disgrams
#library(VennDiagram)
```

<!-- resources -->
```{r functions}
### ================
###   R resources
### ================
source(here("functions", "R_all_functions_v2.R"))
```

```{r filepaths}
### ===========================
###   define constants
### ===========================
### minimum reads threshold
min_reads <- 100
### rarefying subsample size 
richness_subsample_size <- 1000
### number of samples threshold to show heatmap on the page
sample_threshold <- 100

### mapping file path
mapping_file_fp <- list.files(here("metadata"), pattern = ".tsv|.txt|.csv", full.names = TRUE)

### otu table file path
feature_table_fp <- here("PIPITS_output", "out_process", "otu_table.txt")
### taxonomic assignment 
taxo_assignment_fp <- here("BLAST_BROCC_output", "out_brocc", "brocc.log")
### quality control
qc_reads_fp <- here("reads", "reads.log")

```

```{r metadata, warning=F}
### ===========================
###   read in data
### ===========================

col.old <- c("subject_id", "host_species")
col.new <- c("SubjectID", "HostSpecies")

### read mapping file
s <- read_qiime_mapping_file(mapping_file_fp) %>%
  filter(rowSums(is.na(.)|.=="") != ncol(.)) %>%
  rename_at(vars(col.old), ~col.new)

### check for the column names to assign color_by and shape_by for pcoa plots
### and check for number of elements in color_by to expand color palette 
color_by <- NULL
shape_by <- NULL
color_number <- NULL
color_wheel <- NULL
potential_headers <- c("study_group", "SampleType", "study_day", "SubjectID",
                       "current_antibiotics", "HostSpecies", "cage_number") #pick 2
header_idx = which(is.element(potential_headers, colnames(s)))
if(length(header_idx)>0) {
  color_by <- potential_headers[header_idx[1]]
  color_number <- s[color_by] %>% unique() %>% nrow()
  color_wheel <- colorRampPalette(brewer.pal(8, "Set2"))(color_number)
}

if(length(header_idx)>1) {
  shape_by <- potential_headers[header_idx[2]]
}

### BROCC taxonomy
brocc <- read.delim(taxo_assignment_fp)

### read otu table
counts <- readr::read_delim(feature_table_fp, skip=1, delim="\t", quote = "") %>%
  mutate_at(vars(-c(`#OTU ID`, taxonomy)), as.numeric) %>%
  column_to_rownames(var = "#OTU ID") %>% 
  select(-c(taxonomy)) %>%
  as.matrix()

### get read counts
read_counts <- colSums(counts) %>% 
  as.data.frame() %>%
  setNames(c("Read_Counts")) %>%
  rownames_to_column(var="SampleID")

### find the samples to keep
s <- merge(s, read_counts, by="SampleID", all.x=T) %>%
  mutate(Keep = ifelse(!is.na(Read_Counts), Read_Counts > min_reads, FALSE)) %>%
  mutate(isControl = grepl("EBneg.*|Extract.*|Vibriolambda.*|Blank.*|MockDNA.*|DNAfreewater.*|geneblock.*", SampleID, ignore.case = TRUE))

### brocc taxonomic assignment
ta <- read_delim(file=taxo_assignment_fp, delim="\t") %>%
  select(Sequence, Classification) %>%
  mutate(Classification = sub("^Eukaryota;", "k__", Classification)) %>%
  mutate(Classification = sub(';','; p__', Classification)) %>%
  mutate(Classification = sub('(;.*?);', '\\1; c__', Classification)) %>%
  mutate(Classification = sub('(;.*?)(;.*?);', '\\1\\2; o__', Classification)) %>%
  mutate(Classification = sub('(;.*?)(;.*?)(;.*?);', '\\1\\2\\3; f__', Classification)) %>%
  mutate(Classification = sub('(;.*?)(;.*?)(;.*?)(;.*?);', '\\1\\2\\3\\4; g__', Classification)) %>%
  mutate(Classification = sub('(;.*?)(;.*?)(;.*?)(;.*?)(;.*?);', '\\1\\2\\3\\4\\5; s__', Classification)) %>%
  arrange(order(match(rownames(counts), `Sequence`)))

### check if the order of the assignments and the order of featue table is the same
if (!all(rownames(counts) == ta$`Sequence`)) {
  stop (simpleError("The order of the features in the table and classifications don't match"))
}
adf <- split_assignments(ta$Classification) 

# ### pipits taxonomic assignment
# ta <- readr::read_delim(feature_table_fp, skip=1, delim="\t", quote = "") %>%
#   select(`#OTU ID`, taxonomy) %>%
#   arrange(order(match(rownames(counts), `#OTU ID`)))
# 
# ### check if the order of the assignments and the order of featue table is the same
# if (!all(rownames(counts) == ta$`#OTU ID`)) {
#   stop (simpleError("The order of the features in the table and classifications don't match"))
# }
# adf <- split_assignments(ta$taxonomy) 

### remove contamination
is_mitochondrial <- grepl("mitochondria", adf$Family)
is_chloroplast <- grepl("Chloroplast", adf$Class)
is_unassigned <- is.na(adf$Phylum)
is_archaea <- grepl("Archaea", adf$Kingdom)
is_contam <- is_mitochondrial | is_chloroplast | is_unassigned ### Archaea kept to check positive control samples
counts <- counts[!is_contam,]
counts <- counts[, colSums(counts) > 0]
adf <- adf[!is_contam,]
ta <- ta[!is_contam,]
rm(is_contam, is_mitochondrial, is_chloroplast, is_unassigned, is_archaea)
a <- simplify_assignments(adf, rank1="Phylum", rank2="Species")
names(a) <- ta$`Sequence`

### conventional relative abundance
summed_cts <- rowsum(counts, a) 
summed_props <- sweep(summed_cts, 2, colSums(summed_cts), "/")
#if we want all the otus
# all_otus <- counts %>%
#   merge(ta[,c("Sequence","Confidence","trunc_taxon")], by.x="row.names", by.y="Sequence", all.x=T) %>%
#   dplyr::rename(Taxon = trunc_taxon) %>%
#   dplyr::rename(`Sequence` = Row.names)
# 
# write.table(x = all_otus, file = here("output", "OTUs", "all_otus.tsv"), sep = "\t", row.names = F, quote = FALSE)

# ###picogreen corrected abundance
# #get picogreen conc
# picoGreen <- s$final_library_conc_ng_ul
# names(picoGreen) <- s$SampleID
# picoGreen <- picoGreen[colnames(counts)]
# 
# #get picogreen otu table
# otu_props <- sweep(counts, 2, colSums(counts), "/")
# otu_props <- sweep(otu_props, 2, picoGreen, "*")
# 
# neg_otu_props <- otu_props[,grepl("EBneg.*|Extract.*|Blank.*|DNAfreewater.*", colnames(otu_props), ignore.case = TRUE)]
# neg_otu_props <- neg_otu_props[rowSums(neg_otu_props) != 0, colSums(neg_otu_props) > 0]
# 
# #get cumulative area under curve neg ctrl otu props
# otu_ecdf <- neg_otu_props %>%
#   melt() %>%
#   filter(value!=0) %>%
#   select(value) %>%
#   as.matrix() %>%
#   ecdf()
# 
# #threshold for filtering out picogreen otu
# ctrl_thresh <- quantile(otu_ecdf, 0.95)
# 
# otu_props[otu_props < ctrl_thresh] = 0
# otu_props <- otu_props[rowSums(otu_props) != 0, colSums(otu_props) > 0]
# otu_a <- a[match(rownames(otu_props), names(a))]
# summed_otu_props <- rowsum(otu_props, otu_a)

```

\newpage
```{r, Samples error check 1}
### ===========================
###   check for missing samples
### ===========================

### possible issue 1: Samples found in the sample sheet but not in the feature table (0 reads)
s_missing <- s %>%
  filter(!SampleID %in% colnames(counts)) %>%
  select(SampleID, SampleType, isControl)

if (any(!s_missing$isControl)) {
  pander(filter(s_missing, !isControl), caption="These samples were in the sample sheet but not in the feature table.")
  #stop (simpleError("Please fix"))
}

s[s$SampleID %in% s_missing$SampleID, "Keep"] <- FALSE
```


```{r, Samples error check 2}
### possible issue 2: Samples found in the feature table but not in the sample sheet. There must be an error!
in_counts_not_in_s <- setdiff(colnames(counts), s$SampleID)
if (length(in_counts_not_in_s) > 0) {
  stop (simpleError("These SampleID(s) are in the feature table, but not found in the sample sheet.", paste(in_counts_not_in_s, collapse=" ")))
}
```

```{r alpha diversity, warning=F}
### ===========================
###   calculate / read in alpha diversity
### ===========================

# dplyr::filter(!isControl) %>%
s <- s %>%
  merge(vegan::diversity(t(counts)), by.x="SampleID", by.y="row.names", all.x=T) %>%
  dplyr::rename(shannon = y) %>%
  merge(rarefy(t(counts), richness_subsample_size), by.x="SampleID", by.y="row.names", all.x=T) %>%
  dplyr::rename(richness = y)
```

```{r investigator}
### ===========================
###   extract investigator and run date from the sample sheet
### ===========================

all_dates <- as.character(unique(s$run_start_date))
run_date <- paste(lapply(all_dates, change_date_format), collapse=', ')
investigator <- paste(unique(s$investigator)[1], collapse = ", ")
```
\newpage
# Introduction
This report is based on the results of sequencing performed on `r run_date` for `r investigator`. 

## Quality control trimming

```{r reads_trim}
qc <- read_delim(qc_reads_fp, delim = "\t", col_names = FALSE) %>%
  dplyr::rename(counts = "X1") %>%
  mutate(filename = gsub(".*/", "", X2)) %>%
  mutate(direction = ifelse(grepl("_R1", filename), "Forward", "Reverse")) %>%
  mutate(SampleID = gsub("_.*", "", filename)) %>%
  mutate(trim_stage = ifelse(grepl("_R[12].fastq", filename), "Initial demultiplex",
                      ifelse(grepl("_R[12]_rf_trimmed.fastq", filename), "Primer trim",
                      ifelse(grepl("_R[12]_cleaned.fastq", filename), "Filtered",
                      ifelse(grepl("_R[12].fasta", filename), "FastA",
                      ifelse(grepl("_R[12]_vsearch_trimmed.fastq", filename), "Vsearch trim",
                      ifelse(grepl("_R[12]_vsearch_trimmed_paired.fastq", filename), "Pairing", NA))))))) %>%
  mutate(trim_stage = factor(trim_stage, levels = c("Initial demultiplex", "Primer trim", "Filtered", "FastA", "Vsearch trim", "Pairing"))) %>%
  group_by(SampleID, direction) %>%
  mutate(initial_reads = ifelse(trim_stage == "Initial demultiplex", counts, NA)) %>%
  fill(initial_reads, .direction = "downup") %>%
  ungroup() %>%
  mutate(perc_reads = ifelse(initial_reads != 0, counts/initial_reads, 0)) %>%
  merge(select(s, SampleID, SampleType), by = "SampleID", all.x = TRUE)

qc %>%
  ggplot(aes(y = perc_reads, x = trim_stage, group = SampleID)) +
    geom_line(aes(color = SampleType)) +
    geom_point(aes(fill = SampleType), shape = 21) +
    facet_wrap(~direction, ncol = 1) +
    scale_y_continuous(labels = scales::percent) +
    labs(y = "Percent of reads kept", x = "Trimming stage") +
    theme(axis.text.x=element_text(angle=-25, hjust= .1))

```
\newpage
## Histogram of high quality paired reads per sample
The black dashed vertical line shows the minimum number of reads (`r min_reads`) for analysis. Control samples, if any, were included in the histogram.

```{r reads_histogram, fig.width=6, fig.height=4, warning=F}
ggplot(s, aes(x=Read_Counts)) +
    geom_histogram(aes(fill=sample_type), bins=100, boundary = min_reads) +
    geom_vline(xintercept = min_reads, color="black", linetype="dashed") +
    theme_classic() +
    theme_bw() + 
    xlab("Number of reads in sample") +
    ylab("Number of samples")
```

\newpage

## Whole samples that are above the `r min_reads` read count threshold

```{r table}
pander(table(s[, color_by], factor(ifelse(s$Keep, "Keep", "Discard"))))
```

## Distribution of final library concentration

Distribution of final library concentration is given below. Control samples were labeled

```{r}

s %>%
  mutate(Read_Counts = ifelse(is.na(Read_Counts), 1E-6, Read_Counts)) %>%
  droplevels() %>%
  ggplot(aes(x = final_library_conc_ng_ul, y = Read_Counts, fill = SampleType)) +
    geom_point(shape = 21) +
    geom_hline(yintercept = min_reads, color="black", linetype="dashed") +
    scale_y_continuous(trans = "log10") +
    guides(fill = guide_legend(override.aes = list(shape = 21))) +
    labs(y = "Read Counts", x = "Final library DNA conc (ng/uL)")

```

<!-- ## PicoGreen-corrected OTU abundances -->

<!-- Histogram of PicoGreen-corrected OTU abundances in control samples. OTUs in all samples are considered contaminants if they fall below the 95% limit of the abundance distribution (left of the red line) and will be ignored -->

<!-- ```{r} -->

<!-- neg_otu_props %>% -->
<!--   melt() %>% -->
<!--   filter(value!=0) %>% -->
<!--   ggplot(aes(x=value)) + -->
<!--     geom_histogram(binwidth = 0.1) + -->
<!--     theme_bw() + -->
<!--     geom_vline(xintercept = ctrl_thresh, color="red", linetype="dashed") + -->
<!--     labs( -->
<!--       x="PicoGreen corrected OTU abundances in negative control samples (ng/uL)", -->
<!--       y="Number of taxa", -->
<!--       title=paste("All OTUs below the PicoGreen corrected OTU abundance threshold of ", round(ctrl_thresh, 2), "ng/uL will be ignored") -->
<!--     ) -->

<!-- ``` -->

\blandscape

# Taxonomic heatmap

```{r heatmap constants}
prop_cut <- 0.05
thre <- 0.7
satu_limit <- 0.4
heatmap_fp <- here("otu_heatmap.pdf")
show.text <- sum(s$Keep) > sample_threshold
```

Each column of the heatmap represents one sample and each row represents one taxon, typically a genus. Taxa were included in the chart if they are present in at least `r 100*thre`% of samples

The chart is colored white if taxa were not observed in the sample, dark blue if taxa were observed at very low abundance. This allows the reader to quickly survey presence/absence. Abundance values exceeding `r 100*satu_limit`% are colored red, indicating an extremely dominant species.

`r if(show.text){paste0("Please see attached plot otu_heatmap.pdf")}`

```{r heatmap, fig.height=12, fig.width=16}
s_toPlot <- s %>%
  filter(Keep)
  
props_toPlot <- summed_props[, s_toPlot$SampleID]
#if groups contain NA, such as for controls, include NA as a factor level
grps <- c(color_by, shape_by)
### grouped heatmap
if (dim(s_toPlot)[1] > sample_threshold) {
  heatmap_grouped(props_toPlot, s_toPlot, grps=grps, option=1, thre = thre, prop_cut = prop_cut, satu_limit=satu_limit, fname = heatmap_fp)
} else {
  heatmap_grouped(props_toPlot, s_toPlot, grps=grps, option=1, thre = thre, prop_cut = prop_cut, satu_limit=satu_limit)
}
```

\elandscape

# Alpha Diversity

Alpha diversity was assessd by the expected number of observed OTUs (out of rarefying sample size of `r richness_subsample_size`) and Shannon index.

## Number of observed OTUs

```{r richness}
alpha_measure <- "richness"
s %>% filter(Keep) %>%
  filter(!isControl) %>%
  ggplot(aes(x=eval(parse(text=color_by)), y=eval(parse(text=alpha_measure)), color=eval(parse(text=color_by)))) +
  geom_boxplot(outlier.alpha = 0) +
  geom_point(size = 2, position = position_jitterdodge()) +
  labs(y=alpha_measure, x=color_by, color=color_by) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=-25, hjust= .1),
        panel.grid = element_blank(),
        strip.background = element_blank()) +
  #guides(color=F) +
  scale_fill_manual(values = color_wheel) +
  labs(x="study group", color="study group")
  #scale_color_brewer(palette = "Set2")
```

## Shannon Index

```{r shannon}
alpha_measure <- "shannon"
s %>% filter(Keep) %>%
  filter(!isControl) %>%
  ggplot(aes(x=eval(parse(text=color_by)), y=eval(parse(text=alpha_measure)), color=eval(parse(text=color_by)))) +
  geom_boxplot(outlier.alpha = 0) +
  geom_point(size = 2, position = position_jitterdodge()) +
  labs(y=alpha_measure, x=color_by, color=color_by) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=-25, hjust= .1),
        panel.grid = element_blank(),
        strip.background = element_blank()) +
  #guides(color=F) +
  scale_fill_manual(values = color_wheel) +
  labs(x="study group", color="study group")
  #scale_color_brewer(palette = "Set2")
```


\newpage 

# Beta diversity

Similarity between samples was assessed by Bray Curtis and Jaccard distances. 

## Bray-Curtis distance

### PCoA plot based on Bray-Curtis distance

Here, we use Bray-Curtis distance to compare the species composition of the samples to each other.

The first plot shows the distance between each pair of samples in a single 2D plot.  It is not possible to plot the distances exactly on paper, so we have used a method of ordination called Principal Coordinates Analysis to select the best coordinate system for display.  The percentage of total variance captured along each axis is displayed on the chart.

```{r bray-curtis, fig.height=5, fig.width=7}
bc <- vegdist(t(summed_props))
dist_in <- usedist::dist_subset(bc, s_toPlot$SampleID)
plot(make_pcoa_plot(dm = dist_in, s = s_toPlot, color_by=color_by, shape_by=shape_by))

#manual
#plot(make_pcoa_plot(dist_in, s_toPlot, color_by="SampleType", shape_by="expected_genus"))
```
\newpage

\newpage

## Jaccard distance

Here, we use Jaccard distance to compare samples based on shared species membership.  Plots are described above.

### PCoA plot based on Jaccard distance

```{r jaccard, fig.height=5, fig.width=7}
jd <- vegdist(t(summed_props), binary=TRUE)

dist_in <- usedist::dist_subset(jd, s_toPlot$SampleID)
plot(make_pcoa_plot(dist_in, s_toPlot,  color_by=color_by, shape_by=shape_by))
```

\newpage

# APPENDIX: Counts of high quality paired reads for each sample
```{r appendix}
s %>% 
  select(SampleID, Read_Counts, final_library_conc_ng_ul, Keep) %>%
  dplyr::rename(library_ng_ul = "final_library_conc_ng_ul") %>%
  arrange(-Read_Counts) %>% 
  pander()
```

